Issues in developing multivariable molecular signatures for guiding clinical care decisions

### 2016-05-18 Documents/paper-revised.tex
incorporated lisas edits, waiting on final discussion revisions

---
title: "Issues in developing multivariable molecular signatures for guiding clinical care decisions"
author: "Michael C Sachs and Lisa M McShane"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
    pdf_document:
      citation_package: natbib
      keep_tex: yes
      fig_caption: yes
      pandoc_args: ["--bibliography", "paper.bib"]
header-includes:
    - \usepackage{setspace}
    - \doublespacing
    - \usepackage{lineno}
    - \linenumbers
fontsize: 11pt
geometry: margin=1in
---

# Abstract

@@

# Introduction

```{r setup, include = FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(survival)
library(ggkm)
library(tidyr)
library(stargazer)

opts_chunk$set(echo = FALSE, message = FALSE, error = FALSE, warning = FALSE)
knit_hooks$set(plot = function(x, options) {
  if ('tikz' %in% options$dev && !options$external) {
    hook_plot_tex(x, options)
  } else hook_plot_md(x, options)
})
```

Omics technologies that generate a large amount of molecular data characterizing patient biospecimens have the potential to provide information about the patients' disease characteristics above and beyond standard clinical and pathological features. By combining the information from a large number of molecular features into a multivariable model, hereafter referred to as a _biomarker signature_, there is the opportunity to identify distinct subgroups of patients for whom clinical management decisions can be personalized. In early-stage disease, for example, a highly prognostic signature might identify a subpopulation of good risk patients that has such a high probability of long survival or disease-free survival that they do not require additional treatment beyond some standard base therapy. Therefore these good risk patients can be spared the risks and side-effects associated with additional therapy. In the context of a specific therapy that targets a particular molecular pathway, a signature may identify a subpopulation of patients that does or does not benefit from that therapy, thereby guiding the decision of whether or not to administer the therapy.

@@

## Terminology and Notation

A **biomarker signature** is a transformation of multiple individual feature measurements, typically molecular characteristics measured on a multiplex assay, to a one-dimensional space. Specifically, let $X$ denote the set of $p$ feature measurements under consideration. The signature is an unknown function $f(X): \mathbb{R}^p \mapsto \mathbb{R}^1$. The signature result may be continuous, take multiple discrete values, or be dichotomous. In principle, the signature could also depend on other clinical or pathological variables in addition to the molecular measurements, but to simplify the discussion we will focus on signatures that depend on molecular feature data only.  Let $S$ denote the development dataset which includes, for each of $n$ represented individuals, a feature vector $X$, an outcome $Y$, and a treatment $Z$.  $S$ is a sample of size $n$ from distribution $\mathcal{P}$ with domain $\mathcal{S}$. Let $\mathcal{F}$ be a mapping from $\mathcal{S}$ to the space of continuous functions, $\mathcal{D}$, with domain $\mathbb{R}^p$ and range $\mathbb{R}$. Thus $\mathcal{F}: \mathcal{S} \mapsto \mathcal{D}$ denotes the process or algorithm through which a particular $f$ is estimated. We do not place any other restrictions on $\mathcal{F}$, it could be a clustering approach, a regression approach, a combination of both, or something else entirely. We will use $\mathcal{F}$ to denote the manner in which $f$ is estimated and will write $f \in \mathcal{F}$ to denote that $f$ is estimated with the class of methods $\mathcal{F}$.

Let $\phi: \mathcal{D} \times \mathcal{S} \mapsto \mathbb{R}$ denote the statistic that quantifies the performance of the function $f$, such as predictive accuracy, mean squared error, or area under the receiver operating characteristic (ROC) curve (AUC). This could also be a measure of association, such as an odds ratio, hazard ratio, or log-rank statistic. This is a function of both $f$ and $S$.  We are interested in estimating $E_\mathcal{P}[\phi_{f^*}(S)]$, which is the expected error under the data generation mechanism (distribution $\mathcal{P}$), for a particular $f^* \in \mathcal{F}$. This allows us to understand how the signature will perform on future observations generated from $\mathcal{P}$. We may also be interested in estimating $E_\mathcal{P}[\phi_f(S)]$ **for all** $f \in \mathcal{F}$, which is the generalization performance for $f$ generated using mechanism $\mathcal{F}$. This doesn't guide outside researchers as to which specific $f$ to use, yet it is useful for development because it tells us how much signal is in the data. As shorthand we will write this as $E_\mathcal{P}[\phi_\mathcal{F}(S)]$.

## Overview of biomarker signature development

The inherent statistical challenge is how to both develop a signature $f \in \mathcal{F}$ and obtain a valid assessment of its performance.  Additionally, one should provide a specification of $f$ for others to use. Typically, a specific $f$ is estimated using $\mathcal{F}$ based on some training data. The class of methods $\mathcal{F}$ can comprise a variety of different computational approaches. In recent years there has been an explosion in the literature of computational approaches to classification and prediction, and we do not intend to summarize them all here. Some excellent reviews are provided by @hastie2009elements and @moons2012riskI. The main considerations in signature estimation are identifying the features to include, deciding what transformations to apply to the feature measurements, determining how to combine the feature measurements, and determining what, if any, thresholds or cutoffs should be applied to the resulting signature value.

Our focus here will be mainly on the development of signatures using supervised methods, meaning methods that use information on an outcome variable it is desired to predict during the signature development process.  Regression methods are a commonly used class of supervised methods.  Oncotype DX [@paik2004multigene] is an example of a prognostic signature used for clinical decision making in early stage breast cancer that was developed using supervised methods.  In the case of Oncotype DX, the outcome $Y$ was time to distant recurrence of breast cancer and the feature data were gene expression values measured in breast tumor specimens.  A common supervised approach to identifying treatment-selection signatures is to use regression techniques to estimate a signature that has a strong interaction with a particular treatment. Supervised methods are contrasted with unsupervised methods which use only the feature data to derive a signature.  An example of a signature developed by unsupervised methods is the signature that identifies biological subclasses of diffuse large B-cell lymphoma, which were originally developed using clustering methods and subsequently were found to be associated with prognosis [@alizadeh2000distinct]. It is possible, and quite common in high-dimensional data settings, to combine multiple approaches when estimated a signature. For instance, a data-reduction step by variable selection or clustering may be performed before doing regression analysis on the resulting components. A review by Subramanian and Simon identified a large collection of gene expression–based prognostic signatures in lung cancer which had been developed using a wide variety of methods [@subramanian2010gene].

The focus of this paper is how to obtain a valid estimate of signature performance as quantified by a metric $phi$, that is, a good estimate of $E_\mathcal{P}[\phi_\mathcal{F}(S)]$. The principles discussed apply no matter what signature development method is used.  Performance depends on the true signal in the data and the specific algorithm $\mathcal{F}$ used for development.  The performance will reflect that expected of signatures developed applying the prescribed methods to data drawn from the distribution $\mathcal{P}$.  For signature deployment, one would also want a particular specification of $f$ for others to use on independent data (e.g., in clinical practice). It is not trivial to conduct this evaluation in a valid manner when there is a limited data set available to both define a signature and assess its performance.  We illustrate the potential for bias to creep into the performance estimation and review some strategies to avoid the pitfalls that lead to these biases.  We evaluate various strategies through a simulation study and compare the performance of some of those strategies using a real data exmaple involving the development of a signature for use in making treatment decisions for patients with lung cancer.

# Issues

The main goal of interest here is to estimate $E_\mathcal{P}[\phi_\mathcal{F}(S)]$, the expected value of a given performance metric on future observations for $f \in \mathcal{F}$. This can be estimated with the in-sample empirical estimate: $\hat{E}[\phi_f(S)] = \frac{1}{n}\sum_{i=1}^n\phi_f(s_i)$ for a particular $f$. However, if $S$ is used to estimate $f$ then the estimate will be biased due to overfitting, that is, $|E_\mathcal{P}[\phi_f(S)] - \hat{E}[\phi_f(S)]|$ will be large. Overfitting occurs when a model is fit to noise in the data. This often occurs when fitting a model that is overly complex relative to the amount of signal in the available data. Overfitting bias results from the fact that $\phi_f(S)$ depends on $f$ which depends on $S$, and thus the statistic $\phi_f(S)$ is being adaptively defined based on the observed data $S$. Any estimate of signature performance is potentially subject to bias if overfitting occurs, but such biases frequently go unrecognized in the medical literature.

Typically a performance metric $\phi$ used during signature development is degined to quantify either calibration or discrimination, or some combination of the two. Calibration assesses how well the signature based predictions compare to the observed outcomes. Discrimination assesses how well the signature distinguishes between groups of patients who do or do not experience an event.  To assess discrimination, a different metric $\phi$ may be used, such as the area under the ROC curve. Measures of association such as the odds ratio, hazard ratio, or difference in survival probabilities are also commonly used performance metrics, although their value for biomarker signatures has been debated [@pepe2004limitations]. In @zhu2010prognostic, $\phi$ was the hazard ratio comparing the high risk and low risk groups. The risk groups were determined by the signature $f$, which was estimated using the JBL.10 observation arm data. The signature was then applied to the same data that were used to build it to define risk groups, producing a the hazard ratio estimate of 15 for a prognostic effect in the observation arm. Subsequent evaluations of the signature's performance on independent data resulted in hazard ratio estimates in the 2 - 3 range, substantially smaller than the original estimate of 15. Next we reanalyze the JBR.10 data and illustrate some methods to avoid bias in evaluation of signature performance.

## Avoiding Overfitting

A traditional method to avoid overfitting is the split sample approach. First, randomly partition $S$ into the training sample $S_t$ and the holdout sample $S_h$ with sample sizes $n_t$ and $n_h$, respectively. Then, $S_h$ is hidden from the analyst while $\mathcal{F}$ is applied to $S_t$ to estimate the signature function $f_t$. For fixed $f_t$, $\hat{E}[\phi_{f_t}(S_h)]$ is an unbiased estimator of $E_{\mathcal{P}}[\phi_{f_t}(S)]$. @dobbin2011optimally investigate how to optimally split a dataset into training and holdout partitions. The specific form of $f_t$ that is fixed using $S_t$ can be reported as the function for others to use, therefore the aforementioned estimator applies for that specific $f_t$. The drawback of the split-sample approach is that the performance of $f_t$ is likely to be inferior to the performance of a function $f^*$ derived using the same approach applied to the entire data set. @dobbin2007sample investigate how to choose the sample size n so that there is high probability that the signature developed will have expected performance within some specified margin of the the optimal signature that could be developed if the development data set had infinite sample size.

Another approach to avoid overfitting is cross-validation, which is a resampling based approach. For a fixed integer $k$, which can be between 1 and $n$, we randomly partition the full data set $S$ into subgroups of size $k$ (assume for simplicity here that $n$ is evenly divisible by $k$ so that $n/k$ subgroups are formed). For each $k$, $f_{-k}$ is estimated and fixed by appling $\mathcal{F}$ on $S_{-k}$ which is the subset of $S$ that is disjoint from $S_k$. Then, we get an estimate $\hat{E}[\phi_{f_{-k}}(S_k)]$ which is an estimate of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$. This process is repeated $K$ times to yield $K$ estimates and then these are averaged to obtain the performance estimate. This process is called "leave $k$ out" or "$n/k$ fold" cross-validation. Note that for each subgroup in the partition, we obtain a new estimate of $f$, therefore we are only estimating $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$ and not the performance for a single specified signature. Typically, if a specific form for $f$ is desired, it would be estimated using the entire dataset $S$ to yield $f^*$ as above. Importantly one should report the cross-validated estimate just described and not the naive esitimate that would be obtained by substituting the full data $S$ into the performance metric calculation for the signature $f^*$.  The problematic aspects of such naive "resubstitution estimates" of performance are discuss in more detail below.

A variation on the cross-validation approach is bootstrapping. In that case, a sample $S_b$ of size $n$ is sampled _with replacement_ from $S$. Then $f_{b}$ is estimated and fixed by applying $\mathcal{F}$ to $S_{b}$. The performance metric $\phi$ is calculated on the subset of $S$ that is disjoint from $S_b$, which we denote by $S_{-b}$, to yield an estimate of $E[\phi_{f_b}(S_{-b})]$. This process is repeated $K$ times to yield $K$ estimates. These $K$ performance metric estimates are averaged to obtain the mean over the bootstrap replicates. @efron1997improvements suggest a variation, the 0.632 estimate: 
$$
\hat{E}^*[\phi_{\mathcal{F}}(S)] = .368 \hat{E}[\phi_{f}(S)] + 0.632 \hat{E}[\phi_{f_b}(S_{-b})],
$$
where $\hat{E}[\phi_{f}(S)]$ is the naive estimate of $\phi_f$ using the entire dataset.

Another variation on all of these methods is the concept of pre-validation [@tibshirani2002pre]. With pre-validation, instead of computing the statistic $\phi$ for each of the held-out subsets ($S_{-b}$ for the bootstrap or $S_{k}$ for cross-validation), the fitted signature $\hat{f}(X_i)$ is estimated for $X_i \in S_{-b}$ where $\hat{f}$ is estimated using $S_{b}$. This process is repeated to obtain a set of pre-validated signature estimates $\hat{f}$ which are then used to calculate $\phi$. For single-step split sample training and validation, this process is equivalent to what is described above. For cross-validation and the bootstrap, this process avoids the problem of having too few cases to estimate the statistic $\phi$ on each of the smaller held-out datasets.

## Simulation Study

@@

To illustrate some of the different properties of these estimates and how they help to avoid overfitting, we conduct a limited simulation study. Data were generated with 1000 observations, each with a binary outcome $Y$ with prevalence 0.3, and 500 mutually independent features sampled from the standard normal distribution. This is the null case where no features are associated with $Y$. The signature development procedure entails a feature selection step, in which each feature is regressed against $Y$ in a univariate logistic regression model. The 25 features with the smallest p-values are selected for inclusion in a multivariable logistic regression model which defines the final signature.

We compare each of the methods described above: split-sample holdout, cross-validation, bootstrap, and pre-validation, along with several commonly used but biased approaches. The biased approaches are described in Table \ref{descript}. Two of the biased approaches use the full sample to select the features, followed by fitting the multivariable model on the holdout subset. This is referred to as "parital holdout" or "partial CV" when using split-sample holdout or cross-validation as the validation step, respectively. We also implemented the naïve resubstitution approach, wherein the model is trained and evaluated on the same dataset, and the partial resubstitution approach wherein the model is developed on a training subset but then evaluated on the combined training adn holdout data sets. Our main interest is in comparing the bias and variance of the resulting estimates of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$. In our simulation, we look at two different performance metrics, the area under the ROC curve (AUC) and the odds ratio for the outcome comparing the signature groups.

```{r cvsims, fig.cap = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) in the setting where a dataset is used to both develop the signature and evaluate its performance. The violin plots show mirrored density estimates for the AUC for 1000 replicates of the numerical experiment. In each replicate, there are 1000 observations and 500 features. The true value of the AUC is 0.5. CV = Cross validation. \\label{fig1}"}
load("../Code/cvsim-result.RData")
cvres <- do.call("rbind", cvsims)
cvlong <- do.call("rbind", lapply(1:ncol(cvres), function(i){

  cl <- cvres[, i]
  splow <- strsplit(colnames(cvres)[i], ".", fixed = TRUE)
  nm <- paste(unlist(sapply(splow, function(s) rev(rev(s)[-1]))), collapse = ".")
  cls <- sapply(splow, function(s) rev(s)[1])
  data.frame(value = cvres[, i], stat = cls, scen = nm, stringsAsFactors = FALSE)

}))
cvlong$value[cvlong$stat == "OR"] <- exp(cvlong$value[cvlong$stat == "OR"])

cvlong$scen <- factor(cvlong$scen, levels = c("naive", "zhu.cv", "zhu.hold", "zhu.hold2", 
                                              "cv.preval", "cv.10", "cv.100", "holdout.3", 
                                              "holdout.5", "boot"), ordered = TRUE)

labs <- c("Resubstitution", "Partial CV", 
  "Partial Holdout", "Partial Resubstitution", 
  "Pre-validation", "Leave 10 out CV", "Leave 100 out CV", 
  "30% Holdout", "50% Holdout", "Bootstrap")

names(labs) <- levels(cvlong$scen)

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{longtable,booktabs}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Michael C Sachs and Lisa M McShane},
            pdftitle={Issues in developing multivariable molecular signatures for guiding clinical care decisions},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Issues in developing multivariable molecular signatures for guiding
clinical care decisions}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Michael C Sachs and Lisa M McShane}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{May 18, 2016}

\usepackage{setspace}
\doublespacing
\usepackage{lineno}
\linenumbers

\begin{document}

\maketitle

\section{Abstract}\label{abstract}

@@

\section{Introduction}\label{introduction}

Omics technologies that generate a large amount of molecular data
characterizing patient biospecimens have the potential to provide
information about the patients' disease characteristics above and beyond
standard clinical and pathological features. By combining the
information from a large number of molecular features into a
multivariable model, hereafter referred to as a \emph{biomarker
signature}, there is the opportunity to identify distinct subgroups of
patients for whom clinical management decisions can be personalized. In
early-stage disease, for example, a highly prognostic signature might
identify a subpopulation of good risk patients that has such a high
probability of long survival or disease-free survival that they do not
require additional treatment beyond some standard base therapy.
Therefore these good risk patients can be spared the risks and
side-effects associated with additional therapy. In the context of a
specific therapy that targets a particular molecular pathway, a
signature may identify a subpopulation of patients that does or does not
benefit from that therapy, thereby guiding the decision of whether or
not to administer the therapy.

@@

\subsection{Terminology and Notation}\label{terminology-and-notation}

A \textbf{biomarker signature} is a transformation of multiple
individual feature measurements, typically molecular characteristics
measured on a multiplex assay, to a one-dimensional space. Specifically,
let \(X\) denote the set of \(p\) feature measurements under
consideration. The signature is an unknown function
\(f(X): \mathbb{R}^p \mapsto \mathbb{R}^1\). The signature result may be
continuous, take multiple discrete values, or be dichotomous. In
principle, the signature could also depend on other clinical or
pathological variables in addition to the molecular measurements, but to
simplify the discussion we will focus on signatures that depend on
molecular feature data only. Let \(S\) denote the development dataset
which includes, for each of \(n\) represented individuals, a feature
vector \(X\), an outcome \(Y\), and a treatment \(Z\). \(S\) is a sample
of size \(n\) from distribution \(\mathcal{P}\) with domain
\(\mathcal{S}\). Let \(\mathcal{F}\) be a mapping from \(\mathcal{S}\)
to the space of continuous functions, \(\mathcal{D}\), with domain
\(\mathbb{R}^p\) and range \(\mathbb{R}\). Thus
\(\mathcal{F}: \mathcal{S} \mapsto \mathcal{D}\) denotes the process or
algorithm through which a particular \(f\) is estimated. We do not place
any other restrictions on \(\mathcal{F}\), it could be a clustering
approach, a regression approach, a combination of both, or something
else entirely. We will use \(\mathcal{F}\) to denote the manner in which
\(f\) is estimated and will write \(f \in \mathcal{F}\) to denote that
\(f\) is estimated with the class of methods \(\mathcal{F}\).

Let \(\phi: \mathcal{D} \times \mathcal{S} \mapsto \mathbb{R}\) denote
the statistic that quantifies the performance of the function \(f\),
such as predictive accuracy, mean squared error, or area under the
receiver operating characteristic (ROC) curve (AUC). This could also be
a measure of association, such as an odds ratio, hazard ratio, or
log-rank statistic. This is a function of both \(f\) and \(S\). We are
interested in estimating \(E_\mathcal{P}[\phi_{f^*}(S)]\), which is the
expected error under the data generation mechanism (distribution
\(\mathcal{P}\)), for a particular \(f^* \in \mathcal{F}\). This allows
us to understand how the signature will perform on future observations
generated from \(\mathcal{P}\). We may also be interested in estimating
\(E_\mathcal{P}[\phi_f(S)]\) \textbf{for all} \(f \in \mathcal{F}\),
which is the generalization performance for \(f\) generated using
mechanism \(\mathcal{F}\). This doesn't guide outside researchers as to
which specific \(f\) to use, yet it is useful for development because it
tells us how much signal is in the data. As shorthand we will write this
as \(E_\mathcal{P}[\phi_\mathcal{F}(S)]\).

\subsection{Overview of biomarker signature
development}\label{overview-of-biomarker-signature-development}

The inherent statistical challenge is how to both develop a signature
\(f \in \mathcal{F}\) and obtain a valid assessment of its performance.
Additionally, one should provide a specification of \(f\) for others to
use. Typically, a specific \(f\) is estimated using \(\mathcal{F}\)
based on some training data. The class of methods \(\mathcal{F}\) can
comprise a variety of different computational approaches. In recent
years there has been an explosion in the literature of computational
approaches to classification and prediction, and we do not intend to
summarize them all here. Some excellent reviews are provided by
\citet{hastie2009elements} and \citet{moons2012riskI}. The main
considerations in signature estimation are identifying the features to
include, deciding what transformations to apply to the feature
measurements, determining how to combine the feature measurements, and
determining what, if any, thresholds or cutoffs should be applied to the
resulting signature value.

Our focus here will be mainly on the development of signatures using
supervised methods, meaning methods that use information on an outcome
variable it is desired to predict during the signature development
process. Regression methods are a commonly used class of supervised
methods. Oncotype DX \citep{paik2004multigene} is an example of a
prognostic signature used for clinical decision making in early stage
breast cancer that was developed using supervised methods. In the case
of Oncotype DX, the outcome \(Y\) was time to distant recurrence of
breast cancer and the feature data were gene expression values measured
in breast tumor specimens. A common supervised approach to identifying
treatment-selection signatures is to use regression techniques to
estimate a signature that has a strong interaction with a particular
treatment. Supervised methods are contrasted with unsupervised methods
which use only the feature data to derive a signature. An example of a
signature developed by unsupervised methods is the signature that
identifies biological subclasses of diffuse large B-cell lymphoma, which
were originally developed using clustering methods and subsequently were
found to be associated with prognosis \citep{alizadeh2000distinct}. It
is possible, and quite common in high-dimensional data settings, to
combine multiple approaches when estimated a signature. For instance, a
data-reduction step by variable selection or clustering may be performed
before doing regression analysis on the resulting components. A review
by Subramanian and Simon identified a large collection of gene
expression--based prognostic signatures in lung cancer which had been
developed using a wide variety of methods \citep{subramanian2010gene}.

The focus of this paper is how to obtain a valid estimate of signature
performance as quantified by a metric \(phi\), that is, a good estimate
of \(E_\mathcal{P}[\phi_\mathcal{F}(S)]\). The principles discussed
apply no matter what signature development method is used. Performance
depends on the true signal in the data and the specific algorithm
\(\mathcal{F}\) used for development. The performance will reflect that
expected of signatures developed applying the prescribed methods to data
drawn from the distribution \(\mathcal{P}\). For signature deployment,
one would also want a particular specification of \(f\) for others to
use on independent data (e.g., in clinical practice). It is not trivial
to conduct this evaluation in a valid manner when there is a limited
data set available to both define a signature and assess its
performance. We illustrate the potential for bias to creep into the
performance estimation and review some strategies to avoid the pitfalls
that lead to these biases. We evaluate various strategies through a
simulation study and compare the performance of some of those strategies
using a real data exmaple involving the development of a signature for
use in making treatment decisions for patients with lung cancer.

\section{Issues}\label{issues}

The main goal of interest here is to estimate
\(E_\mathcal{P}[\phi_\mathcal{F}(S)]\), the expected value of a given
performance metric on future observations for \(f \in \mathcal{F}\).
This can be estimated with the in-sample empirical estimate:
\(\hat{E}[\phi_f(S)] = \frac{1}{n}\sum_{i=1}^n\phi_f(s_i)\) for a
particular \(f\). However, if \(S\) is used to estimate \(f\) then the
estimate will be biased due to overfitting, that is,
\(|E_\mathcal{P}[\phi_f(S)] - \hat{E}[\phi_f(S)]|\) will be large.
Overfitting occurs when a model is fit to noise in the data. This often
occurs when fitting a model that is overly complex relative to the
amount of signal in the available data. Overfitting bias results from
the fact that \(\phi_f(S)\) depends on \(f\) which depends on \(S\), and
thus the statistic \(\phi_f(S)\) is being adaptively defined based on
the observed data \(S\). Any estimate of signature performance is
potentially subject to bias if overfitting occurs, but such biases
frequently go unrecognized in the medical literature.

Typically a performance metric \(\phi\) used during signature
development is degined to quantify either calibration or discrimination,
or some combination of the two. Calibration assesses how well the
signature based predictions compare to the observed outcomes.
Discrimination assesses how well the signature distinguishes between
groups of patients who do or do not experience an event. To assess
discrimination, a different metric \(\phi\) may be used, such as the
area under the ROC curve. Measures of association such as the odds
ratio, hazard ratio, or difference in survival probabilities are also
commonly used performance metrics, although their value for biomarker
signatures has been debated \citep{pepe2004limitations}. In
\citet{zhu2010prognostic}, \(\phi\) was the hazard ratio comparing the
high risk and low risk groups. The risk groups were determined by the
signature \(f\), which was estimated using the JBL.10 observation arm
data. The signature was then applied to the same data that were used to
build it to define risk groups, producing a the hazard ratio estimate of
15 for a prognostic effect in the observation arm. Subsequent
evaluations of the signature's performance on independent data resulted
in hazard ratio estimates in the 2 - 3 range, substantially smaller than
the original estimate of 15. Next we reanalyze the JBR.10 data and
illustrate some methods to avoid bias in evaluation of signature
performance.

\subsection{Avoiding Overfitting}\label{avoiding-overfitting}

A traditional method to avoid overfitting is the split sample approach.
First, randomly partition \(S\) into the training sample \(S_t\) and the
holdout sample \(S_h\) with sample sizes \(n_t\) and \(n_h\),
respectively. Then, \(S_h\) is hidden from the analyst while
\(\mathcal{F}\) is applied to \(S_t\) to estimate the signature function
\(f_t\). For fixed \(f_t\), \(\hat{E}[\phi_{f_t}(S_h)]\) is an unbiased
estimator of \(E_{\mathcal{P}}[\phi_{f_t}(S)]\).
\citet{dobbin2011optimally} investigate how to optimally split a dataset
into training and holdout partitions. The specific form of \(f_t\) that
is fixed using \(S_t\) can be reported as the function for others to
use, therefore the aforementioned estimator applies for that specific
\(f_t\). The drawback of the split-sample approach is that the
performance of \(f_t\) is likely to be inferior to the performance of a
function \(f^*\) derived using the same approach applied to the entire
data set. \citet{dobbin2007sample} investigate how to choose the sample
size n so that there is high probability that the signature developed
will have expected performance within some specified margin of the the
optimal signature that could be developed if the development data set
had infinite sample size.

Another approach to avoid overfitting is cross-validation, which is a
resampling based approach. For a fixed integer \(k\), which can be
between 1 and \(n\), we randomly partition the full data set \(S\) into
subgroups of size \(k\) (assume for simplicity here that \(n\) is evenly
divisible by \(k\) so that \(n/k\) subgroups are formed). For each
\(k\), \(f_{-k}\) is estimated and fixed by appling \(\mathcal{F}\) on
\(S_{-k}\) which is the subset of \(S\) that is disjoint from \(S_k\).
Then, we get an estimate \(\hat{E}[\phi_{f_{-k}}(S_k)]\) which is an
estimate of \(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\). This process is
repeated \(K\) times to yield \(K\) estimates and then these are
averaged to obtain the performance estimate. This process is called
``leave \(k\) out'' or ``\(n/k\) fold'' cross-validation. Note that for
each subgroup in the partition, we obtain a new estimate of \(f\),
therefore we are only estimating
\(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\) and not the performance for a
single specified signature. Typically, if a specific form for \(f\) is
desired, it would be estimated using the entire dataset \(S\) to yield
\(f^*\) as above. Importantly one should report the cross-validated
estimate just described and not the naive esitimate that would be
obtained by substituting the full data \(S\) into the performance metric
calculation for the signature \(f^*\). The problematic aspects of such
naive ``resubstitution estimates'' of performance are discuss in more
detail below.

A variation on the cross-validation approach is bootstrapping. In that
case, a sample \(S_b\) of size \(n\) is sampled \emph{with replacement}
from \(S\). Then \(f_{b}\) is estimated and fixed by applying
\(\mathcal{F}\) to \(S_{b}\). The performance metric \(\phi\) is
calculated on the subset of \(S\) that is disjoint from \(S_b\), which
we denote by \(S_{-b}\), to yield an estimate of
\(E[\phi_{f_b}(S_{-b})]\). This process is repeated \(K\) times to yield
\(K\) estimates. These \(K\) performance metric estimates are averaged
to obtain the mean over the bootstrap replicates.
\citet{efron1997improvements} suggest a variation, the 0.632 estimate:
\[
\hat{E}^*[\phi_{\mathcal{F}}(S)] = .368 \hat{E}[\phi_{f}(S)] + 0.632 \hat{E}[\phi_{f_b}(S_{-b})],
\] where \(\hat{E}[\phi_{f}(S)]\) is the naive estimate of \(\phi_f\)
using the entire dataset.

Another variation on all of these methods is the concept of
pre-validation \citep{tibshirani2002pre}. With pre-validation, instead
of computing the statistic \(\phi\) for each of the held-out subsets
(\(S_{-b}\) for the bootstrap or \(S_{k}\) for cross-validation), the
fitted signature \(\hat{f}(X_i)\) is estimated for \(X_i \in S_{-b}\)
where \(\hat{f}\) is estimated using \(S_{b}\). This process is repeated
to obtain a set of pre-validated signature estimates \(\hat{f}\) which
are then used to calculate \(\phi\). For single-step split sample
training and validation, this process is equivalent to what is described
above. For cross-validation and the bootstrap, this process avoids the
problem of having too few cases to estimate the statistic \(\phi\) on
each of the smaller held-out datasets.

\subsection{Simulation Study}\label{simulation-study}

@@

To illustrate some of the different properties of these estimates and
how they help to avoid overfitting, we conduct a limited simulation
study. Data were generated with 1000 observations, each with a binary
outcome \(Y\) with prevalence 0.3, and 500 mutually independent features
sampled from the standard normal distribution. This is the null case
where no features are associated with \(Y\). The signature development
procedure entails a feature selection step, in which each feature is
regressed against \(Y\) in a univariate logistic regression model. The
25 features with the smallest p-values are selected for inclusion in a
multivariable logistic regression model which defines the final
signature.

We compare each of the methods described above: split-sample holdout,
cross-validation, bootstrap, and pre-validation, along with several
commonly used but biased approaches. The biased approaches are described
in Table \ref{descript}. Two of the biased approaches use the full
sample to select the features, followed by fitting the multivariable
model on the holdout subset. This is referred to as ``parital holdout''
or ``partial CV'' when using split-sample holdout or cross-validation as
the validation step, respectively. We also implemented the naïve
resubstitution approach, wherein the model is trained and evaluated on
the same dataset, and the partial resubstitution approach wherein the
model is developed on a training subset but then evaluated on the
combined training adn holdout data sets. Our main interest is in
comparing the bias and variance of the resulting estimates of
\(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\). In our simulation, we look
at two different performance metrics, the area under the ROC curve (AUC)
and the odds ratio for the outcome comparing the signature groups.

\begin{figure}[htbp]
\centering
\includegraphics{paper-revised_files/figure-latex/cvsims-1.pdf}
\caption{Comparison of different approaches to estimating the Area Under
the ROC Curve (AUC) in the setting where a dataset is used to both
develop the signature and evaluate its performance. The violin plots
show mirrored density estimates for the AUC for 1000 replicates of the
numerical experiment. In each replicate, there are 1000 observations and
500 features. The true value of the AUC is 0.5. CV = Cross validation.
\label{fig1}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{paper-revised_files/figure-latex/cvsims2-1.pdf}
\caption{Comparison of different approaches to estimating the odds ratio
(OR) in the setting where a dataset is used to both develop the
signature and evaluate its performance. The violin plots show mirrored
density estimates for the log OR for 1000 replicates of the numerical
experiment. In each replicate, there are 1000 observations and 500
features. The true value of the OR is 1.0. CV = Cross validation.
\label{fig2}}
\end{figure}
